Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-02-25 article-models-tt.tex
expanding intro of section 3

\section{Type theory without type constructors}\label{tt}
In this appendix we give a possible description of type theory without basic
constructors and with explicit extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't lists of variable
declarations. The reasons for this are twofold: (1) in the models of type theory
such as a univalent universe or the graph model, contexts are types and graphs
respectively -- they are objects rather than lists of variable declarations; we
intend to give a general algebraic description of higher categories. Besides,
we will not have to be burdened with comments about variables being bounded or not, or fresh or
not occuring at all.

@@

Much of the rules we state are just compatibility rules of extension, weakening
and substitution with each other. To get the overview, we list the sections
where these compatibility rules are described in the following table (in this
table, the
subsection mentioned in row $X$ and column $Y$ consideres the rules of the
operation $Y\circ X$):

\begin{center}
\begin{tabular}{r|ccc}
& extension & weakening & substitution\\
\hline
extension & \autoref{comp-ee} & \autoref{comp-ew} & \\
weakening & \autoref{comp-we} & \autoref{comp-ww} & \autoref{comp-ws}\\
substitution & \autoref{comp-se} & \autoref{comp-sw} & \autoref{comp-ss}
\end{tabular}
\end{center}

Since extension by a context acts trivially on families and terms there are
no rules added for extension followed by substitution.

>>>>

\section{Type theory before type constructors}\label{tt}
In this section we give a description of type theory before basic
constructors and with explicit extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't lists of variable
declarations. The reasons for this are twofold: (1) in the models of type theory
such as a univalent universe or the graph model, contexts are types and graphs
respectively -- they are objects rather than lists of variable declarations; we
intend to give a general algebraic description of higher categories. Besides,
we will not have to be burdened with comments about variables being bounded or not, or fresh or
not occuring at all.

We will develop a theory which is in fact a theory of contexts and families of
contexts where a family of context can be a type, behind which we have the
intuition that it is an atomic or irreducible family of contexts. We get immediately
that every construction can be made in an arbitrary context.

@@

Much of the rules we state are just compatibility rules of extension, weakening
and substitution with each other. In a way, these rules assert that our contexts
are just structureless lists of contexts and that likewise terms are structureless
lists of terms. They are structureless in the sense that the order in which
they are formed by pairing up is irrelevant. We note that this causes complications
in the traditional way that categorical sematics of type theory is implemented,
where contexts become objects of the category which is supposed to model type
theory. The reason for this is that context extension will not satisfy all the
compatibility rules we're about to state. The first step to resolving this is taking
the types in the empty context as the objects.

To get the overview of the compatibility rules, we list the sections
where these compatibility rules are described in the following table (in this
table, the
subsection mentioned in row $X$ and column $Y$ consideres the rules of the
operation $Y\circ X$):
\begin{center}
\begin{tabular}{r|ccc}
& extension & weakening & substitution\\
\hline
extension & \autoref{comp-ee} & \autoref{comp-ew} & \\
weakening & \autoref{comp-we} & \autoref{comp-ww} & \autoref{comp-ws}\\
substitution & \autoref{comp-se} & \autoref{comp-sw} & \autoref{comp-ss}
\end{tabular}
\end{center}
Since extension by a context acts trivially on families and terms there are
no rules added for extension followed by substitution.
